---
title: "valueR - helping characterize uncertainty in value modeling"
output: github_document
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
This R package can be installed with devtools::install_github("caddell/valueR")

Please review the associated paper (https://doi.org/10.1002/mcda.1717) for details and cite as appropriate.
```{r message=FALSE, warning=FALSE}
#install from github if needed
#devtools::install_github("caddell/valueR")
library(valueR)
#cite using citation
citation("valueR")
library(tidyverse)
library(ggthemes)
```
Build a set of alternatives for us to investigate.

```{r message=FALSE, warning=FALSE}
set.seed(123)
n = 1000

orange = tibble(
  alternative = c(rep("orange", n)),
  value = c(rnorm(n, mean= 67, sd=3)),
  cost = c(rnorm(n, mean = 59, sd= 4))
)

green = tibble(
  alternative = c(rep("green", n)),
  value = c(rnorm(n, mean= 75, sd=3)),
  cost = c(rnorm(n, mean = 69, sd= 3))
)

blue = tibble(
  alternative = c(rep("blue", n)),
  value = c(rnorm(n, mean= 63, sd=2)),
  cost = c(rnorm(n, mean = 75, sd= 2))
)

yellow = tibble(
  alternative = c(rep("yellow", n)),
  value = c(rnorm(n, mean= 71, sd=1)),
  cost = c(rnorm(n, mean = 67, sd=1))
)

purple = tibble(
  alternative = c(rep("purple", n)),
  value = c(rnorm(n, mean= 74, sd=1.5)),
  cost = c(rnorm(n, mean = 72.5, sd=1.25))
)

red = tibble(
  alternative = c(rep("red", n)),
  value = c(rnorm(n, mean= 80, sd=1.5)),
  cost = c(rnorm(n, mean = 70, sd=2.25))
)

#bind all alternatives together for future use
all.alternatives = rbind(orange, green, blue, yellow, purple, red)

```

## Plot the data

The first step to any problem: Visualize the data. This is how we normally see the output of a simulation.

```{r message=FALSE, warning=FALSE}
#plot alternative as mean of alternatives
mean.alt = all.alternatives %>% group_by(alternative) %>% summarise_all(mean)

#plot means of alternatives
ggplot(data = mean.alt, aes(x = cost, y = value, color = alternative, fill = alternative))+
  theme_gdocs()+
  geom_point(size = 4)+
  labs(x = "Cost", y = "Value", title = "Mean Alternative Outcomes", color= "Alternative")+
  guides(fill = FALSE, alpha = FALSE)+
  coord_fixed(xlim = c(45,80), ylim = c(55,85))+
  theme(plot.background=element_blank())+
  scale_color_manual(values = c("blue", "green", "orange", "purple","red","yellow"))



```

## Plot ALL the data

Try it again with all the alternatives and densities.
```{r message=FALSE, warning=FALSE}
#plot alternative as points w/ densities
ggplot(data = all.alternatives, aes(x = cost, y = value, color = alternative, fill = alternative))+
  theme_gdocs()+
  stat_density2d(aes(alpha=..level.., color = alternative), bins = 20, geom = "polygon")+
  geom_point(alpha = .1)+
  labs(x = "Cost", y = "Value", title = "Uncertain Alternative Outcomes", fill= "Alternative")+
  guides(color = FALSE, alpha = FALSE)+
  coord_fixed()+
  theme(plot.background=element_blank())+
  scale_fill_manual(values = c("blue", "green", "orange", "purple","red","yellow"))+
  scale_color_manual(values = c("blue", "green", "orange", "purple","red","yellow"))
```

This visualization helps us better understand the spread and uncertainty of an alternative. It gives the stakeholder more information. Unfortunately, this can be unhelpful at times and lead to decision paralysis. What do we do next?

## Level 1 Anlaysis

We've found it helpful to talk about alternative pairwise comparisons in terms of dominance. In general, you have four types of dominance. An alternative is dominate over another if the alternative returns more value at the same or less cost. Naturally it is, dominated if it returns less value for more cost. If an alternative returns more value, but for more cost, then we consider it pareto-optimal plus. If it has less value, but costs less, we consider it pareto-optimal minus.

If we consdier the means of our alternatives, we can easily identify dominance. Lets consider red and blue.
```{r message=FALSE, warning=FALSE}
#plot means of alternatives
ggplot(data = mean.alt %>% filter(alternative == "blue" | alternative == "red"), aes(x = cost, y = value, color = alternative, fill = alternative))+
  theme_gdocs()+
  geom_point(size = 4)+
  labs(x = "Cost", y = "Value", title = "Mean Alternative Outcomes", color= "Alternative")+
  guides(fill = FALSE, alpha = FALSE)+
  coord_fixed(xlim = c(45,80), ylim = c(55,85))+
  theme(plot.background=element_blank())+
  scale_color_manual(values = c("blue","red"))
```

As we can see, red returns more value for less cost and is therefore dominate. But if we consider all the data we understand it isn't that simple. 

```{r message=FALSE, warning=FALSE}
#plot alternative as points w/ densities
ggplot(data = all.alternatives %>% filter(alternative == "red" | alternative == "blue"), aes(x = cost, y = value, color = alternative, fill = alternative))+
  theme_gdocs()+
  stat_density2d(aes(alpha=..level.., color = alternative),geom = "polygon")+
  geom_point(alpha = .1)+
  labs(x = "Cost", y = "Value", title = "Uncertain Alternative Outcomes", fill= "Alternative")+
  guides(color = FALSE, alpha = FALSE)+
  coord_fixed()+
  theme(plot.background=element_blank())+
  scale_fill_manual(values = c("blue", "red"))+
  scale_color_manual(values = c("blue", "red"))
```

In this case, we've decided to compare all possible combinations. We call this Level 1 Analysis. 
```{r message=FALSE, warning=FALSE}
level1_red_blue = level1(red, blue, 500)
level1_red_blue$final_table
```

From this analysis we can tell that red is a more dominate alternative. In fact, we can tell you that it will dominate blue over 95% of the time. This adds another layer of understanding for a decision maker. They probably didn't need it for this simple of a solution, but this style of analysis lends it self to more complicated decisions. 
```{r message=FALSE, warning=FALSE}
#plot alternative as points w/ densities
ggplot(data = all.alternatives %>% filter(alternative == "red" | alternative == "green"), aes(x = cost, y = value, color = alternative, fill = alternative))+
  theme_gdocs()+
  stat_density2d(aes(alpha=..level.., color = alternative),geom = "polygon")+
  geom_point(alpha = .1)+
  labs(x = "Cost", y = "Value", title = "Uncertain Alternative Outcomes", fill= "Alternative")+
  guides(color = FALSE, alpha = FALSE)+
  coord_fixed()+
  theme(plot.background=element_blank())+
  scale_fill_manual(values = c("green", "red"))+
  scale_color_manual(values = c("green", "red"))
```
This plot is much more interesting, but it would be helpful to quantify the alternatives. 
```{r message=FALSE, warning=FALSE}
level1_green_red = level1(green,red)
level1_green_red$final_table
```

## Level 2
Lets suppose that our decision maker is comfortable "picking" the red alternative. We should try and give them an expectation of this choice. 

```{r message=FALSE, warning=FALSE}
expectations = level2(green,red, .25)
expectations$trade

```
So how happy are we?
```{r message=FALSE, warning=FALSE}
expectations$hist
```







